Do motif roles provide unique information about species risk of extinction?\\ \medskip Supplementary Information

### 2021-10-01 manuscript/supplement_motifs.tex
added Annas raw text for persistence calc

    \subsection{Stochastic extinction}

@@

        As well as the intrinsic baseline probability of extinction, extinction risk for consumers is affected by access to resources. 
		When all resources are extant, the probability of extinction for a consumer $i$, $P(\lnot i)$, is equal to the baseline probability; $P(\lnot i|X_{1},...,X_{n}) = \pi_{base}$. 
		When all resources are extinct, the consumer will go extinct; $P(\lnot i|\lnot X_{1},...,\lnot X_{n})=1$.

		Calculating $P(i)$ or $P(\lnot i)$ is less straightforward when a consumer $i$ loses a fraction $k/n = f$ of its prey. 
		Plausible ecological responses to losing some prey include i) a topological response, where the consumer's probability of extinction is constant unless all resources are lost, ii) a linear response, in which the extinction probability increases linearly with resources loss, and iii) several nonlinear responses where the probability of extinction increases non-linearly with the fraction resource lost. 
		The extinction probability will differ depending on the response function. 
		We use a sigmoid non-linear functional response of consumers to the loss of resources ($\alpha >1, \beta >1$) which has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}. 
		We use a cumulative density function of a beta distribution for this purpose:
		\begin{equation}
		\label{betafunc}
		P(\lnot i|f) = \pi_{i} + (1 - \pi_{i}) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}
		\end{equation}

    \subsection{Numerical simulation}

>>>>

        We use Bayesian Network modelling to calculate the persistence of each species in our simulated networks. 
        A Bayesian Network represents a graphical model for probabilistic relationships among a set of variables. The nodes of the network represents the variables, and the links represent conditional dependencies among the variables. Each node has a set of local probability dependencies, conditional on the state of the parent nodes. 
        In our ecological setting, species act as nodes and are Bernoulli random variables, being either extant or extinct. The links act as feeding interactions between the nodes.
        The extinction probability of each species is a function of the state of its resources.

    \subsection{Baseline probability of extinction}

        Consider a consumer species with several resource species $n$. 
        The extinction probability
        of this consumer will depend on both the total state of its resource species $n$, and on a
        baseline extinction probability of extinction, $\pi_{base}$. 
        $\pi_{base}$ represents the risk of going extinct
        given factors not related to the food web itself, and could reflect diseases or stochastic
        extinctions of small populations. 
        All the species in the network retain a baseline
        probability of extinction, which can be different to reflect greater sensitivity in some groups of species.

    \subsection{Extinction due to food-web effects}

        If a consumer with $n$ resources loses $k$ of them, the fraction of resources lost $f$ will be
        $f = (k/n)$. 
        Note that we here make a strong assumption, where we only care about the fraction of resources lost and not the identity of the resources species.

@@

        The response of a consumer to the loss of a fraction of its resources can be modelled in several biologically plausible ways. For example, the response could be linear or non-linear.
        A sigmoid non-linear functional response of consumers to the loss of resources ($\alpha >1, \beta >1$) has been shown to most accurately capture the secondary extinctions produced by a dynamical model~\citep{Eklof2013}.
        Therefore, we use a cumulative density function of a beta distribution to get each species \textit{i} probability of extinction:
        \begin{equation}
        \label{betafunc}
        P(\lnot i|f) = \pi_{base} + (1 - \pi_{base}) \frac{B(f;\alpha,\beta)}{B(\alpha,\beta)}
        \end{equation}

        If all resource species of a consumer \textit{C} are present ($f = 0$), the extinction probability will be $P(\lnot C|f) = \pi_{base}$. 
        Similarly, if all resources are extinct ($f = 1$), the extinction probability $P(\lnot C|f)$ will be equal to 1 and the consumer will go extinct.
        If the fraction $f$ is not 1 nor 0, the non-linear sigmoid curve will determine the response of the consumer, which will be more severe if the fraction resources lost is high and less severe if the fraction is small. This will in turn affect the extinction probability $P(\lnot C|f)$. 
        For a basal species, we assume that abiotic resources are always available. Therefore, a primary producer is independent of resource species and will always have $f = 0$.

        % how we calculate persistences 
        As consumer species depends on its resource species, we start determining the status of each primary producer in the network.
        Systematically, we use eqn \ref{betafunc} to receive each species probability of extinction $P(\lnot i|f)$, and next we perform Bernoulli trials to determine whether the species goes extinct or not. 
        Imagine a basal species \textit{A} with extinction probability $P(\lnot A) = \pi_{base}$ where $\pi_{base} = 0.1$. We draw a random number $r$ from a uniform distribution between $0-1$. The basal species would be considered extinct if $r < 0.1$, and extant otherwise. 
        We use this information when performing a Bernoulli trial on a consumer \textit{C} connected to this basal species, inserting our information of the $\pi_{base}$ and fraction of resources lost $f$ of this consumer, into eqn \ref{betafunc} to 
        receive $P(\lnot C|f)$. In the same manner as for the basal species above, we compare $P(\lnot C|f)$ with a random number $r$, and if $r < P(\lnot C|f)$ the consumer species is considered extinct. 
        In this way, we evaluate each species existence, starting from the bottom and working our way up.
        As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
        For each species, we then estimate the fraction of simulations in which the species persist, and this value is the approximated probability of persistence.
        To approximate persistence probabilities through numerical simulation are shown to be highly efficient and produce the same result as exact methods of solving Bayesian networks \citep{Haussler2020}.

    % various disturbance scenarios
        In our baseline scenario, primary producers had a baseline extinction probability of $\pi_{base} = 0.1$. 
        We also simulated a set of scenarios where primary producers had a higher baseline probability of extinction ($\pi_{disturbed}$). 
        These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
        $\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. 
        The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
        Consumer species retained $\pi_{base}=0.1$ in all cases.
